A tool set for random number generation on GPUs in R(?)

### 2021-11-01 pkg/gpuRandom/inst/documents/paper1_2021_5/paper1.Rnw
removed unnecessary cpp files
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/diseasemapping@2943 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\author{Ruoyong Xu\\University of Toronto
   \And Patrick Brown\\University of Toronto}
\Plainauthor{Ruoyong Xu, Patrick Brown}

\title{Random number generation and applications on GPU’s in R(?)}
\Plaintitle{}
\Shorttitle{}

\Abstract{
Parallel processing by graphics processing unit (GPU) can be used to speed up computationally intensive tasks, which when combined with \proglang{R}, it can largely improve \proglang{R}’s downsides in terms of slow speed, memory usage and computation mode.  There is currently no \proglang{R} package that does random number generation on GPU, while random number generation is critical in simulation-based statistical inference and modeling. We introduce the \proglang{R} package \pkg{gpuRandom} that leverages the \pkg{gpuR} package, and can generate random numbers on GPU by utilizing the \pkg{clRNG} (OpenCL) library, the package enables reproducible research by setting random streams on GPU and can thus accelerate several types of simulation and modelling.  \pkg{gpuRandom} is portable and flexible, developers can use its random number generation kernel for various other purposes and applications. 
}

@@

Random number generation is critical in simulation-based statistical inference, machine learning and many other scientific fields. While most random number generators are sequential, the \proglang{R} packages \pkg{parallel} (not on cran, how to cite?) and \pkg{future} \citep*{future1.19.1} are able to generate random numbers in parallel on multicore CPUs. More specifically, \pkg{parallel} writes an interface for the \pkg{RngStreams}, a \proglang{\texttt{C++}} library by \cite{l2002object} which is based on a combined multiple-recursive generator (MRG), \pkg{future} also uses the combined MRG algorithm for generating random numbers, however, it is not very clear how the streams are implemented (?).

@@

\pkg{clRNG} is an \proglang{OpenCL} library for uniform random number generation, it provides four different RNGs: the MRG31k3p, MRG32k3a, LFSR113, and Philox-4×32-10 generators. \pkg{gpuRandom} package uses the MRG31k3p RNG from the \proglang{clRNG}, making it able to generate random numbers on GPUs. In what follows, we will illustrate how to create streams and how to use streams to generate uniform and (standard) Gaussian random numbers.

\subsection{Creating streams}\label{create_streams}
The popular way of obtaining multiple streams is to take an RNG with a long period and cut the RNG's sequence into very long pieces, these pieces are adjacent and of equal length $Z$, so each piece can be used as a separate stream. Creating a new stream amounts to computing its starting point (seed). Each of these streams can also be partitioned into substreams with equally-spaced starting points \citep{l2002object}, which are occasionally useful.

@@

In \proglang{clRNG} library, the initial state of the first stream has a default value (12345). For the MRG31k3p RNG, the entire period length of approximately $2^{185}$ is divided into approximately $2^{51}$ non-overlapping streams of length $Z = 2^{134}$. Each stream is further partitioned into substreams of length $W = 2^{72}$. The state (and seed) of each stream is a vector of six 31-bit integers. This size of state is appropriate for having streams running in work items on GPU cards, for example, while providing a sufficient period length for most applications.

@@

as.vector(gpuRandom::runif(n=10, streams=myStreamsGpu2, Nglobal=c(1,nrow(myStreamsGpu1)), type = "double"))[-(1:6)]
@
The arguments of the \fct{gpuRandom::runif} are described as follows:
\begin{itemize}
  \item \code{n}: a vector of length 2 specifying the row and column number if to create a matrix, or a number specifying the length if to create a vector.
  \item \code{streams}: streams for random number generation.
  \item \code{Nglobal}: global index space. Default is set as $(64,8)$.
  \item \code{type}: ``double'' or ``float'' or ``integer'' format of generated random numbers.
\end{itemize}
\fct{runif} returns the same sequence of random numbers if it uses the same streams in the same current state (here \code{myStreamsGpu1} and \code{myStreamsGpu2}), not streams starting from the same states. Reusing \code{myStreamsGpu1} will produce a vector different from \code{sim_1}. Because each time a random number is generated, the current state of the stream advances by one position. Below shows the \code{myStreamsGpu1} after it being used.
<<eval=TRUE,tidy=TRUE, include=TRUE>>=
t(matrix(as.matrix(myStreamsGpu1), nrow(myStreamsCpu), ncol(myStreamsCpu), dimnames = dimnames(myStreamsCpu)))
@
Unlike the objects created by \proglang{R}, by default they remain in the memory after a restart. Streams created on GPU does not remain in the memory when a new \proglang{R} session begins. However, we can save the streams on the CPU in a file using the \fct{saveRDS} function, and later recall the streams with the \fct{readRDS} function, in this way we can reproduce the exact same sequence of random numbers in simulations. Below is a trivial example, in which we demonstrate saving streams to a data file called \code{streams_from_GPU.rds} on CPU and then load it back and transfer it to GPU.
<<save streams,echo=TRUE,eval=TRUE>>=
saveRDS(as.matrix(createStreamsGpu(n=4)), "myStreams.rds")
# Load the streams object as streams_saved
streams_saved <- vclMatrix(readRDS("myStreams.rds"))
@

@@

random_u<-as.vector(runif(n=50000, streams=streams, Nglobal=c(64,8), type="double"))
@
\begin{figure}[H]
\centering
<<largenumberuniforms, eval=TRUE, echo=FALSE, fig.height=4, fig.width=4, fig.show='hold'>>=
hist(random_u, prob = TRUE, main="Histogram")
plot(ecdf(random_u), ylab="F(x)",main="Empirical cumulative distribution")
curve(punif(x, 0, 1), lty=2, add=TRUE, col="red")
@
\caption{Histogram and empirical cumulative distribution of the GPU-generated random numbers.\label{fig0}}
\end{figure}

@@

In the following code, We create a $1024 \times 512$ matrix %named \code{normal_matrix} 
of single-precision floating-point Gaussian random numbers, and assess Normality by making a histogram and a Q-Q plot of the generated random numbers in Figure \ref{fig2}. The points in the Q-Q plot shown in the right panel of Figure \ref{fig2} fall along the diagonal line, we have good evidence the generated data are from Gaussian distribution. We find doing a Q-Q plot can occupy a large amount of time in the process of producing an \proglang{R} markdown file, so as an incidental work, we adapted the \fct{stat::qnorm} to make the function \fct{gpuRandom::qqnorm} that does the quantile computation for the normal distribution part on a GPU device, which can speed up the Q-Q plot.
<<histAndQQ_prepare,eval=TRUE,cache=TRUE>>=
streams <- createStreamsGpu(n =512*128)
normal_matrix<-gpuRandom::rnorm(c(1024,512), streams=streams, 
                                Nglobal=c(512,128), type="float")
avector<-as.vector(as.matrix(normal_matrix))
@
\begin{figure}[H]
\centering
 <<histAndQQ, echo=TRUE, eval=TRUE, cache=TRUE, fig.height=4, fig.width=4, fig.show='hold'>>=
hist(avector,breaks=40, main = "histogram")
gpuRandom::qqnorm(avector, Nglobal=c(128,64), main="Q-Q Plot")
@
\caption{Histogram and Q-Q plot of the GPU-generated random numbers.\label{fig2}}
\end{figure}

@@

\section{Monte Carlo simulation for Fisher's exact test}
One application of GPU-generated random numbers in \pkg{gpuRandom} is Monte Carlo simulation for Fisher's exact test. Fisher’s exact test is applied for analyzing usually $2 \times 2$ contingency tables when one of the expected values in table is less than 5. Different from methods which rely on approximation, Fisher's exact test computes directly the probability of obtaining each possible combination of the data for the same row and column totals (marginal totals) as the observed table, and get the exact p-value by adding together all the probabilities of tables as extreme or more extreme than the one observed. However, when the observed table gets large in terms of sample size and table dimensions, the number of combinations of cell frequencies with the same marginal totals gets very large, \cite[][p. 23]{mehta2011ibm} shows a $5 \times 6$ observed table that has 1.6 billion possible tables. Calculating the exact P-values may lead to very long run-time and can sometimes exceed the memory limits of your computer. Hence, the option \code{simulate.p.value = TRUE} in \fct{stats::fisher.test} is provided, which enables computing p-values by Monte Carlo simulation for tables larger than 2 $\times$ 2. Given an observed table and a number of replicates $B$, the Monte Carlo simulation for Fisher's exact test does the following steps: 
\begin{enumerate}
  \item Calculate the test statistic for the observed table.
	\item In each iteration, simulate a random table with the same dimensions and marginal totals as the observed table, compute and sometimes save the test statistic from the random table.
	\item Count the number of iterations that have test statistics less or equal to the one from the observed table (\emph{Counts}).
	\item Estimate p-value using $\frac{1 + \emph{Counts}}{B + 1}$.
\end{enumerate}
The test statistic calculated for each random table is $-\sum_{i,j}\log(n_{ij}!), i=(1,\dots, I), j=(1,\dots,J)$, (i.e., minus log-factorial of table), where $I$ and $J$ are the row and column number of the observed table. This test statistic can also be independently calculated for a table by \fct{gpuRandom::logfactSum}.

\fct{stats::fisher.test} can take a long running time when $B$ reaches millions as the $B$ random tables are constructed sequentially, one in each iteration. \fct{gpuRandom::fisher.sim} adapted the backend function \fct{rcont2} used by \fct{stats::fisher.test} for constructing random tables, so that it generates \code{prod(Nglobal)} random tables in parallel in each iteration (i.e., one random table each work-item per iteration), and the steps 2 and 3 
are also done in parallel by work-items at each iteration on GPU. Step 2 of saving test statistics from the random tables is made optional, which can reduce the run-time.

@@

Figure \ref{fig4} visualizes the approximate sampling distributions of the test statistics from the two examples "Month" and "Week", by plotting a histogram for each of the example. The values of observed statistics are calculated by \fct{gpuRandom::logfactSum} and are marked with a blue line on each plot. 
\begin{figure}[H]
\centering
<<fighistMonth, eval=TRUE, echo=FALSE, fig.show='hold', cache=TRUE>>=
hist(month_stats, xlab=TeX('-$\\sum(\\log(n_{ij}!))$'), breaks=40, ylab="Frequency", main="Month data")
abline(v = observed_m, col = "blue", lwd = 1.5)

@@

%Due to the importance of Gaussian random field, there are a large amount of studies devoted to Gaussian random field generation techniques. 
\cite{LiuandLi2019} gives a comprehensive review on 7 popular methods for Gaussian random field generation, which are the turning bands method \citep{matheron_1973}, spectral method \citep{Meja1974OnTS,1972JSV}, matrix decomposition method \citep{davis1987}, Karhunen-Lo\'eve expansion \citep{LoeveM:1978}, moving average method \citep{journel1974, oliver1995moving}, sequential simulation \citep{johnson1987multivariate, gomez1993joint, pebesma2004multivariable}, and local average subdivision \citep{fenton1990simulation}. All of these methods except matrix decomposition method, are approximations and has specific requirements on the type of grid or covariance functions. The matrix decomposition method is exact, it works for all covariance functions and can generate random field on all types of grids. Other \proglang{R} packages that offer simulation of Gaussian random fields like \pkg{geoR} \citep{geoR2001}, does not work for large number of locations; the \pkg{RandomFields} \citep{RandomFields2015,RandomFields2020} package can use different methods for simulation of Gaussian fields, among which the (modified) circulant embedding method \citep{Dietrich1997FastAE} for covariance matrix decomposition is also an exact method (need to confirm???), however, it works only for isotropic Gaussian fields and on rectangular grids. \pkg{gpuRandom} does exact simulation of exact Gaussian fields on the GPU as it relies on the matrix decomposition method. The implementation of this method is as follows.

@@

myRaster = raster::raster( raster::extent(0,Ngrid[1]/Ngrid[2],5,6),Ngrid[1], Ngrid[2])
coordsSp = sp::SpatialPoints(raster::xyFromCell(myRaster, 1:raster::ncell(myRaster)))
coordsGpu = vclMatrix(coordsSp@coords, 
                      nrow(coordsSp@coords), 
                      ncol(coordsSp@coords), type=theType)
@
Step 2, create 5 parameter sets as a small example, in practical studies, there can be hundreds or thousands of parameter sets. Then, convert the matrix of parameters to a ``vclMatrix'' for later use by \fct{gpuRandom::maternGpuParam}.
<<setup paramsBatch, eval=TRUE,echo=FALSE>>=
myParamsBatch = 
rbind(c(shape=1.25, range=0.5, variance = 1.5, nugget = 0,anisoRatio = 1, anisoAngleRadians = 0), 
c(shape=2.15, range=0.25, variance = 2, nugget = 0, anisoRatio = 4, anisoAngleRadians = pi/7),
c(shape=0.55, range=1.5, variance = 2, nugget = 0, anisoRatio = 4, anisoAngleRadians = pi/7),
c(shape=2.15, range=0.5, variance = 2, nugget = 0, anisoRatio = 4, anisoAngleRadians = -pi/7),
c(shape=2.15, range=0.5, variance = 2, nugget = 0, anisoRatio = 2, anisoAngleRadians = pi/4))
@
<<show parameterBatch,eval=TRUE>>=
myParamsBatch
paramsGpu = gpuRandom::maternGpuParam(myParamsBatch, type=theType)
@
Step 3, compute Mat\'ern covariance matrices using \fct{gpuRandom::maternBatch}, the returned Mat\'ern covariance matrices %$\Sigma_1, \Sigma_2, \dots, \Sigma_5$ correspond to the 5 parameter sets respectively, the matrices 
are each of size $4800 \times 4800$ and are stacked by row in the output \code{maternCov}. 
<<r outputBatchCreate, eval=TRUE>>=
maternCov = vclMatrix(0, nrow(paramsGpu)*nrow(coordsGpu), 
                         nrow(coordsGpu),type=theType)
dim(maternCov)
gpuRandom::maternBatch(maternCov, coordsGpu, paramsGpu,
                       Nglobal=c(128,64), Nlocal=c(16,4))
@
Step 4, the first argument \code{maternCov} in \fct{gpuRandom::cholBatch} specifies the matrix object to take Cholesky decomposition. Unit lower triangular matrices $L_i$'s are returned and stacked by row in \code{maternCov}, \code{diagMat} stores the diagonal values of each $D_i$ in a row, for example, if each batch $\Sigma_i$ is of size $n \times n$, then each batch $L_i$ is $n \times n$, and each batch $D_i$ is $1 \times n$.
<<r cholBach, eval=TRUE>>=
diagMat = vclMatrix(0, nrow(paramsGpu), ncol(maternCov), type = theType)
gpuRandom::cholBatch(maternCov, diagMat, numbatchD=nrow(myParamsBatch),
                     Nglobal= c(128, 8),
                     Nlocal= c(32, 8),
                     NlocalCache=1000)
@
% \begin{gather}
% \begin{bmatrix} \Sigma_{1} \\ \Sigma_2 \\ \Sigma_3 \\ \vdots
% \end{bmatrix}
%  \rightarrow
%  \begin{bmatrix}
%   L_{1} \\ L_{2} \\L_{3} \\ \vdots
%   \end{bmatrix} \text{and}
%   \begin{bmatrix}
%   D_{1} \\ D_{2} \\D_{3} \\ \vdots
%   \end{bmatrix}
% \end{gather}
Step 5, generate 2 standard Gaussian random vectors \code{zmatGpu}$=(Z_1, Z_2)$ using \fct{gpuRandom::rnorm}, in which \code{c(nrow(materCov),2)} specifies the number of rows and columns of \code{zmatGpu}.
<<r randomNormalsSim1, eval=TRUE, cache=FALSE>>=
streamsGpu <- createStreamsGpu(n=128*64)
@
<<r randomNormalsSim2, eval=TRUE, cache=FALSE>>=
zmatGpu = rnorm(c(nrow(maternCov),2), streams=streamsGpu, Nglobal=c(128,64),
                 type = theType)
@
Step 6, use \fct{gpuRandom::multiplyLowerDiagonalBatch} to compute $U = L * D^{(1/2)}* Z$ in batches, see the following illustration. \code{simMat} is the output matrix for $U$, \code{maternCov}, \code{diagMat}, and \code{zmatGpu} correspond to the matrices of $L$, $D$ and $Z=(Z_1, Z_2)$ respectively.
<<r multLowerdiag_LDZ, eval=TRUE, cache=FALSE, results='hide'>>=
simMat = vclMatrix(0, nrow(zmatGpu), ncol(zmatGpu), 
                   type = theType)

>>>>

\author{Ruoyong Xu\\University of Toronto
   \And Patrick Brown\\University of Toronto
   \And Pierre L’Ecuyer\\University of Montreal}
\Plainauthor{Ruoyong Xu, Patrick Brown, Pierre L’Ecuyer}

\title{A tool set for random number generation on GPUs in R(?)}
\Plaintitle{}
\Shorttitle{}

\Abstract{
We introduce the \proglang{R} package \pkg{gpuRandom} that leverages the \pkg{gpuR} package, and can generate random numbers on GPU by utilizing the \pkg{clRNG} (OpenCL) library. Parallel processing by graphics processing unit (GPU) can be used to speed up computationally intensive tasks, which when combined with \proglang{R}, it can largely improve \proglang{R}’s downsides in terms of slow speed, memory usage and computation mode.  There is currently no \proglang{R} package that does random number generation on GPU, while random number generation is critical in simulation-based statistical inference and modeling. The package enables reproducible research by setting random streams on GPU and can thus accelerate several types of simulation and modelling.  \pkg{gpuRandom} is portable and flexible, developers can use its random number generation kernel for various other purposes and applications. 
}

@@

Random number generation is critical in simulation-based statistical inference, machine learning and many other scientific fields. While most random number generators are sequential, the \proglang{R} packages \pkg{parallel} (not on cran, how to cite?) and \pkg{future} \citep*{future1.19.1} are able to generate random numbers in parallel on multicore CPUs. More specifically, \pkg{parallel} writes an interface for the \pkg{RngStreams}, a \proglang{\texttt{C++}} library by \cite{l2002object} which is based on a combined multiple-recursive generator (MRG). \pkg{future} also uses the combined MRG algorithm for generating random numbers, however, it is not very clear how the streams are implemented (?).

@@

\pkg{clRNG} is an \proglang{OpenCL} library for uniform random number generation, it provides four different RNGs: the MRG31k3p, MRG32k3a, LFSR113, and Philox-4×32-10 generators. These four RNGs use different types of constructions.  But we can create multiple streams with any of those four types. \pkg{gpuRandom} package uses the MRG31k3p RNG from the \proglang{clRNG}, making it able to generate random numbers on GPUs. In what follows, we will illustrate how to create streams and how to use streams to generate uniform and (standard) Gaussian random numbers.

\subsection{Creating streams}\label{create_streams}
The popular way of obtaining multiple streams is to take an RNG with a long period and cut the RNG's sequence into very long pieces, these pieces are adjacent and of equal length $Z$, so each piece can be used as a separate stream. Creating a new stream amounts to computing its starting point (seed). Each of these streams can also be partitioned into substreams with equally-spaced starting points \citep{l2002object,rLEC15a}, which are occasionally useful.

@@

In \proglang{clRNG} library, for the MRG31k3p RNG, the entire period length of approximately $2^{185}$ is divided into approximately $2^{51}$ non-overlapping streams of length $Z = 2^{134}$. Each stream is further partitioned into substreams of length $W = 2^{72}$. The state (and seed) of each stream is a vector of six 31-bit integers. This size of state is appropriate for having streams running in work items on GPU cards, for example, while providing a sufficient period length for most applications. The initial state of the first stream for the MRG31k3p is by default a vector of six integers ``12345''.

@@

as.vector(gpuRandom::runif(n=10, streams=myStreamsGpu2, Nglobal=c(1,nrow(myStreamsGpu1)), type = "double"))[-(1:6)]
@
The arguments of the \fct{gpuRandom::runif} are described as follows:
\begin{itemize}
  \item \code{n}: a vector of length 2 specifying the row and column number if to create a matrix, or a number specifying the length if to create a vector.
  \item \code{streams}: streams for random number generation.
  \item \code{Nglobal}: global index space. Default is set as $(64,8)$.
  \item \code{type}: ``double'' or ``float'' or ``integer'' format of generated random numbers.
\end{itemize}
\fct{runif} returns the same sequence of random numbers if it uses the same streams in the same current state (here \code{myStreamsGpu1} and \code{myStreamsGpu2}), not streams starting from the same states. Reusing \code{myStreamsGpu1} will produce a vector different from \code{sim_1}. Because each time a random number is generated, the current state of the stream advances by one position. %Below shows the \code{myStreamsGpu1} after it being used.
<<eval=TRUE,tidy=TRUE, include=FALSE>>=
t(matrix(as.matrix(myStreamsGpu1), nrow(myStreamsCpu), ncol(myStreamsCpu), dimnames = dimnames(myStreamsCpu)))
@
Unlike the objects created by \proglang{R}, by default they remain in the memory after a restart. Streams created on GPU does not remain in the memory when a new \proglang{R} session begins. However, we can save the streams on the CPU in a file using the \fct{saveRDS} function, and later recall the streams with the \fct{readRDS} function, in this way we can reproduce the exact same sequence of random numbers in simulations. Below is a trivial example, in which we demonstrate saving streams to a data file called \code{streams_from_GPU.rds} on CPU and then load it back and transfer it to GPU.
<<save streams,echo=TRUE,eval=TRUE>>=
saveRDS(as.matrix(createStreamsGpu(n=4)), "myStreams.rds")
# Load the streams object as streams_saved
streams_saved <- vclMatrix(readRDS("myStreams.rds"))
@

@@

random_u<-as.vector(runif(n=50000, streams=streams, Nglobal=c(64,8), type="double"))
@
\begin{figure}[H]
\centering
<<largenumberuniforms, eval=TRUE, echo=FALSE, fig.height=4, fig.width=4, fig.show='hold', dev='png'>>=
hist(random_u, prob = TRUE, main="Histogram")
plot(ecdf(random_u), ylab="F(x)",main="Empirical cumulative distribution")
curve(punif(x, 0, 1), lty=2, add=TRUE, col="red")
@
\caption{Histogram and empirical cumulative distribution of the GPU-generated random numbers.\label{fig0}}
\end{figure}

@@

In the following code, We create a $1024 \times 512$ matrix %named \code{normal_matrix} 
of single-precision floating-point Gaussian random numbers, and assess Normality by making a histogram and a Q-Q plot of the generated random numbers in Figure \ref{fig2}. The points in the Q-Q plot shown in the right panel of Figure \ref{fig2} fall along the diagonal line, we have good evidence the generated data are from Gaussian distribution. We find doing a Q-Q plot can occupy a large amount of time in the process of producing an \proglang{R} markdown file, so as an incidental work, we adapted the \fct{stat::qnorm} to make the function \fct{gpuRandom::qqnorm} that does the quantile computation for the normal distribution part on a GPU device, which can speed up the Q-Q plot.
<<histAndQQ_prepare,eval=TRUE,cache=TRUE>>=
streams <- createStreamsGpu(n =512*128)
normal_matrix<-gpuRandom::rnorm(c(1024,512), streams=streams, 
                                Nglobal=c(512,128), type="float")
avector<-as.vector(as.matrix(normal_matrix))
@
\begin{figure}[H]
\centering
 <<histAndQQ, echo=TRUE, eval=TRUE, cache=TRUE, fig.height=4, fig.width=4, fig.show='hold',dev='png'>>=
hist(avector,breaks=40, main = "histogram")
gpuRandom::qqnorm(avector, Nglobal=c(128,64), main="Q-Q Plot")
@
\caption{Histogram and Q-Q plot of the GPU-generated random numbers.\label{fig2}}
\end{figure}

@@

\section{Monte Carlo simulation for Fisher's exact test}
One application of GPU-generated random numbers in \pkg{gpuRandom} is Monte Carlo simulation for Fisher's exact test. Fisher’s exact test is applied for analyzing usually $2 \times 2$ contingency tables when one of the expected values in table is less than 5. Different from methods which rely on approximation, Fisher's exact test computes directly the probability of obtaining each possible combination of the data for the same row and column totals (marginal totals) as the observed table, and get the exact p-value by adding together all the probabilities of tables as extreme or more extreme than the one observed. However, when the observed table gets large in terms of sample size and table dimensions, the number of combinations of cell frequencies with the same marginal totals gets very large, \cite[][p. 23]{mehta2011ibm} shows a $5 \times 6$ observed table that has 1.6 billion possible tables. Calculating the exact P-values may lead to very long run-time and can sometimes exceed the memory limits of your computer. Hence, the option \code{simulate.p.value = TRUE} in \fct{stats::fisher.test} is provided, which enables computing p-values by Monte Carlo simulation for tables larger than 2 $\times$ 2.

The test statistic calculated for each random table is $-\sum_{i,j}\log(n_{ij}!), i=(1,\dots, I), j=(1,\dots,J)$, (i.e., minus log-factorial of table), where $I$ and $J$ are the row and column number of the observed table. This test statistic can also be independently calculated for a table by \fct{gpuRandom::logfactSum}. Given an observed table and a number of replicates $B$, the Monte Carlo simulation for Fisher's exact test does the following steps: 
\begin{enumerate}
  \item Calculate the test statistic for the observed table.
	\item In each iteration, simulate a random table with the same dimensions and marginal totals as the observed table, compute and sometimes save the test statistic from the random table.
	\item Count the number of iterations that have test statistics less or equal to the one from the observed table (\emph{Counts}).
	\item Estimate p-value using $\frac{1 + \emph{Counts}}{B + 1}$.
\end{enumerate}

In \fct{stats::fisher.test}, the $B$ random tables are constructed sequentially, one in each iteration. \fct{gpuRandom::fisher.sim} adapted the backend function \fct{rcont2} used by \fct{stats::fisher.test} for constructing random tables, so that it generates \code{prod(Nglobal)} random tables in parallel in each iteration (i.e., one random table each work-item per iteration).  Each random table requires multiple calls to \fct{clrngMrg31k3pNextState} in \fct{rcont2} to generate $U(0,1)$ random numbers, the number of calls depends on the row and column number of the observed table. For more about the algorithm used the function \fct{rcont2} for generating random two-way table with given marginal totals, see \citep{patefield1981efficient}. Step 1 and step 2-3 are done on a GPU with two kernels enqueued sequentially, the test statistics (if returned) and the count? are saved on a GPU, the last step for calculating the p-value is done in R. Step 2 of saving test statistics from the random tables is made optional, which can reduce the run-time. p-value is calculated using the simple random sampling method.

@@

Figure \ref{fig4} visualizes the approximate sampling distributions of the test statistics from the two examples "Month" and "Week", by plotting a histogram for each of the example. The values of observed statistics are calculated by \fct{gpuRandom::logfactSum} and are marked with a blue line on each plot. 
\begin{figure}[H]
\centering
<<fighistMonth, eval=TRUE, echo=FALSE, fig.show='hold', cache=TRUE, dev='png'>>=
hist(month_stats, xlab=TeX('-$\\sum(\\log(n_{ij}!))$'), breaks=40, ylab="Frequency", main="Month data")
abline(v = observed_m, col = "blue", lwd = 1.5)

@@

%Due to the importance of Gaussian random field, there are a large amount of studies devoted to Gaussian random field generation techniques. 
\cite{LiuandLi2019} gives a comprehensive review on 7 popular methods for Gaussian random field generation, %which are the turning bands method \citep{matheron_1973}, spectral method \citep{Meja1974OnTS,1972JSV}, matrix decomposition method \citep{davis1987}, Karhunen-Lo\'eve expansion \citep{LoeveM:1978}, moving average method \citep{journel1974, oliver1995moving}, sequential simulation \citep{johnson1987multivariate, gomez1993joint, pebesma2004multivariable}, and local average subdivision \citep{fenton1990simulation}. 
all of these methods except matrix decomposition method, are approximations and has specific requirements on the type of grid or covariance functions. The matrix decomposition method is exact, it works for all covariance functions and can generate random field on all types of grids. Other \proglang{R} packages that offer simulation of Gaussian random fields like \pkg{geoR} \citep{geoR2001}, does not work for large number of locations; the \pkg{RandomFields} \citep{RandomFields2015,RandomFields2020} package can use different methods for simulation of Gaussian fields, among which the (modified) circulant embedding method \citep{Dietrich1997FastAE} for covariance matrix decomposition is also an exact method (need to confirm???), however, it works only for isotropic Gaussian fields and on rectangular grids. \pkg{gpuRandom} does exact simulation of exact Gaussian fields on the GPU as it relies on the matrix decomposition method. The implementation of this method is as follows.

@@

myRaster = raster::raster( raster::extent(0,Ngrid[1]/Ngrid[2],5,6),Ngrid[1], Ngrid[2])
coordsSp = sp::SpatialPoints(raster::xyFromCell(myRaster, 1:raster::ncell(myRaster)))
coordsGpu = vclMatrix(coordsSp@coords, 
                      nrow(coordsSp@coords), 
                      ncol(coordsSp@coords), type=theType)
@
Step 2, create 5 parameter sets as a small example, in practical studies, there can be hundreds or thousands of parameter sets. Then, convert the matrix of parameters to a ``vclMatrix'' for later use by \fct{gpuRandom::maternGpuParam}.
<<setup paramsBatch, eval=TRUE,echo=FALSE>>=
myParamsBatch = 
rbind(c(shape=1.25, range=0.5, variance = 1.5, nugget = 0,anisoRatio = 1, anisoAngleRadians = 0), 
c(shape=2.15, range=0.25, variance = 2, nugget = 0, anisoRatio = 4, anisoAngleRadians = pi/7),
c(shape=0.55, range=1.5, variance = 2, nugget = 0, anisoRatio = 4, anisoAngleRadians = pi/7),
c(shape=2.15, range=0.5, variance = 2, nugget = 0, anisoRatio = 4, anisoAngleRadians = -pi/7),
c(shape=2.15, range=0.5, variance = 2, nugget = 0, anisoRatio = 2, anisoAngleRadians = pi/4))
@
<<show parameterBatch,eval=TRUE>>=
myParamsBatch
paramsGpu = gpuRandom::maternGpuParam(myParamsBatch, type=theType)
@
Step 3, compute Mat\'ern covariance matrices using \fct{gpuRandom::maternBatch}, the returned Mat\'ern covariance matrices %$\Sigma_1, \Sigma_2, \dots, \Sigma_5$ correspond to the 5 parameter sets respectively, the matrices 
are each of size $4800 \times 4800$ and are stacked by row in the output \code{maternCov}. 
<<r outputBatchCreate, eval=TRUE>>=
maternCov = vclMatrix(0, nrow(paramsGpu)*nrow(coordsGpu), 
                         nrow(coordsGpu),type=theType)
dim(maternCov)
gpuRandom::maternBatch(maternCov, coordsGpu, paramsGpu,
                       Nglobal=c(128,64), Nlocal=c(16,4))
@
Step 4, the first argument \code{maternCov} in \fct{gpuRandom::cholBatch} specifies the matrix object to take Cholesky decomposition. Unit lower triangular matrices $L_i$'s are returned and stacked by row in \code{maternCov}, \code{diagMat} stores the diagonal values of each $D_i$ in a row, for example, if each batch $\Sigma_i$ is of size $n \times n$, then each batch $L_i$ is $n \times n$, and each batch $D_i$ is $1 \times n$.
<<r cholBach, eval=TRUE>>=
diagMat = vclMatrix(0, nrow(paramsGpu), ncol(maternCov), type = theType)
gpuRandom::cholBatch(maternCov, diagMat, numbatchD=nrow(myParamsBatch),
                     Nglobal= c(128, 8),
                     Nlocal= c(32, 8),
                     NlocalCache=1000)
@
% \begin{gather}
% \begin{bmatrix} \Sigma_{1} \\ \Sigma_2 \\ \Sigma_3 \\ \vdots
% \end{bmatrix}
%  \rightarrow
%  \begin{bmatrix}
%   L_{1} \\ L_{2} \\L_{3} \\ \vdots
%   \end{bmatrix} \text{and}
%   \begin{bmatrix}
%   D_{1} \\ D_{2} \\D_{3} \\ \vdots
%   \end{bmatrix}
% \end{gather}
Step 5, generate 2 standard Gaussian random vectors \code{zmatGpu}$=(Z_1, Z_2)$ using \fct{gpuRandom::rnorm}, in which \code{c(nrow(materCov),2)} specifies the number of rows and columns of \code{zmatGpu}.
<<r randomNormalsSim1, eval=TRUE, cache=FALSE>>=
streamsGpu <- createStreamsGpu(n=128*64)
@
<<r randomNormalsSim2, eval=TRUE, cache=FALSE>>=
zmatGpu = rnorm(c(nrow(maternCov),2), streams=streamsGpu, Nglobal=c(128,64),
                 type = theType)
@
Step 6, use \fct{gpuRandom::multiplyLowerDiagonalBatch} to compute $U = L * D^{(1/2)}* Z$ in batches, see the following illustration. \code{simMat} is the output matrix for $U$, \code{maternCov}, \code{diagMat}, and \code{zmatGpu} correspond to the matrices of $L$, $D$ and $Z=(Z_1, Z_2)$ respectively.
<<r multLowerdiag_LDZ, eval=TRUE, cache=FALSE, results='hide',dev='png'>>=
simMat = vclMatrix(0, nrow(zmatGpu), ncol(zmatGpu), 
                   type = theType)

gpuRandom::multiplyLowerDiagonalBatch(
  simMat, maternCov, diagMat, zmatGpu,
  diagIsOne = 1L, # diagonal of L is one
  transformD = "sqrt", # take the square root of each element of D
  Nglobal,
  Nlocal,
  NlocalCache)
@
\begin{gather}
 \begin{bmatrix}  L_{1} \\ L_{2} \\L_{3} \\ \vdots
 \end{bmatrix} 
 *
  \begin{bmatrix}
   D_{1} \\ D_{2} \\D_{3} \\ \vdots
   \end{bmatrix} 
   *
   \begin{bmatrix}
   Z_{11} & Z_{12}
   \end{bmatrix}
  =
 \begin{bmatrix}
   L_1D_1Z_{11} & L_1D_1Z_{12} \\
   L_2D_2Z_{11} & L_2D_2Z_{12} \\
   L_3D_3Z_{11} & L_3D_3Z_{12} 
  \end{bmatrix}
\end{gather}
Step 7, Finally, plot the 10 simulated Gaussian random surfaces.
<<r plotSimSetup, echo=TRUE, eval=TRUE, cache=FALSE>>=
library(raster)
simRaster = brick(myRaster, nl = ncol(simMat)*nrow(paramsGpu))
values(simRaster) = as.vector(as.matrix(simMat))
names(simRaster) = apply(expand.grid('par',1:nrow(paramsGpu), 
                                     'sim', 1:ncol(simMat)), 1, paste, collapse='')
@
\begin{figure}[H]
\centering
<<maternplot, echo=TRUE, fig.show='hold', eval=TRUE, dev='pdf',cache=FALSE,dev='png'>>=
par(mar=rep(0.1, 4))
for(D in names(simRaster)) {
  plot(extent(simRaster))
  plot(simRaster[[D]], legend=FALSE, add=TRUE)
}
@
\caption{\label{fig:5} }
\end{figure}
